# Setup

```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(janitor)
library(tidyverse)
```

## Data
Load raw data and clean names, remove NAs where applicable then save csv file for reading into database

```{r}
# Read GDW data
gdw_path <- "data/raw/GDW_v1_0.gdb"
st_layers(gdw_path) # Contains barriers (dams) and reservoirs
gdw <- st_read(gdw_path, layer = "GDW_barriers_v1_0") |> 
    clean_names() |> 
  mutate(across(-Shape, ~ { # Replace -99 in all columns with NA
    if (is.numeric(.)) {
      na_if(., -99)
    } else {
      na_if(., "-99")
    }
  }))
```


# Read BasinATLAS data
basinatlas_path <- "data/raw/BasinATLAS_v10.gdb"
st_layers(basinatlas_path)
basinatlas <- st_read(basinatlas_path, layer = "BasinATLAS_v10_lev12") |> 
    clean_names() # Level 12 because thats what matches GDW

# Read RiverATLAS
riveratlas_path <- "data/raw/RiverATLAS_v10.gdb"
riveratlas <- st_read(riveratlas_path, layer = "RiverATLAS_v10") |> 
    clean_names()

# Add LakeATLAS?

# Read Free-Flowing Rivers
ffr_path <- "data/raw/FFR_river_network.gdb"
ffr <- st_read(ffr_path, layer = "FFR_river_network_v1") |> 
    clean_names() |> 
    rename(hyriv_id = reach_id) # rename reach_id from FFR to hyriv_id used by the rest

```


# Exploration

```{r}
# Looking for future dams GDW

# Find # of GDW dams built in future
gdw %>%
  filter(YEAR_DAM > 2021) %>%
  nrow()

# Plot when dams were built removing -99
gdw |> 
  filter(YEAR_DAM != -99 & YEAR_DAM >= 1900) |> 
  ggplot(aes(x = YEAR_DAM)) +
  geom_bar()

```

### Checking if bas_id (FFR) and hybas_id (basinatlas) are the same 

```{r}
# 1. Check for shared values (Intersection)
shared_ids <- intersect(ffr$bas_id, basinatlas$hybas_id)
length(shared_ids)

# 2. 

ffr %>% slice_sample(n = 5) %>% pull(bas_id) %>% print()
basinatlas %>% slice_sample(n = 5) %>% pull(hybas_id) %>% print()
```


## ICOLD 
```{r}
icold <- readxl::read_excel("data/raw/world_register_dams_2025.xlsx")
1000 dams in ICOLD are under construction
# Same thing for icold
icold |> 
  filter(Year != -99 & Year >= 1900) |> 
  ggplot(aes(x = Year)) +
  geom_bar()
```
